---
title: "In-class_Ex06 : Emerging Hot Spot Analysis"
author: "Xiaohan"
date: "today"
execute:
  eval: true
  echo: true
  freeze: true
---
## 1 Overview

Emerging Hot Spot Analysis (EHSA) is a spatio-temporal statistical technique that identifies and categorises areas where clusters of high or low values of a phenomenon are developing, intensifying, or diminishing over time. It combines the Getis-Ord Gi\* statistic, which measures spatial clustering, with the Mann-Kendall trend test, a non-parametric trend analysis, to detect trends in these clusters within a "space-time cube" of data.

The analysis consists of four main steps:

-   Building a space-time cube,
-   Calculating Getis-Ord local Gi\* statistic for each bin by using an FDR correction,
-   Evaluating these hot and cold spot trends by using Mann-Kendall trend test,
-   Categorising each study area location by referring to the resultant trend z-score and p-value for each location with data, and with the hot spot z-score and p-value for each bin.

## 2 Getting Started

### 2.1 Installing and loading the R packages

`p_load()` of **pacman** package is used to check if the necessary packages have been installed in R; if yes, load them into the R environment.

```{r}
pacman::p_load(sf, sfdep, tmap,
               plotly, tidyverse,
               Kendall)
```

## 3 The Data

Two data sets are used in this in-class exercise:

-   *Hunan*, a geospatial data set in ESRI shapefile format, and
-   *Hunan_GDPPC*, an attribute data set in csv format.

### 3.1 Importing geospatial data

In the code chunk below, `st_read()` of **sf** package is used to import the *Hunan* shapefile into R.

```{r}
hunan <- st_read(
  dsn = "data/geospatial",
  layer = "Hunan") %>%
  st_transform(crs = 32650)
```

### 3.2 Importing attribute table

In the code chunk below, `read_csv()` of **readr** is used to import *Hunan_GDPPC.csv* into R.

```{r}
GDPPC <- read_csv(
  "data/aspatial/Hunan_GDPPC.csv")
```

## 4 Creating a Time Series Cube

In the code chunk below, `spacetime()` of **sfdep** is used to create a spatio-temporal cube.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of **sfdep** package is used to verify if *GDPPC_st* is indeed a space-time cube object.

```{r}
is_spacetime_cube(GDPPC_st)
```

The **TRUE** return confirms that the *GDPPC_st* object is indeed a space-time cube.

## 5 Computing Gi\*

Next, we will compute the local Gi\* statistics.

### 5.1 Deriving the spatial weights

The code chunk below is used to identify neighbours and to derive inverse distance weights.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb,
                             geometry,
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

::: callout-tip
## Things to learn from the code chunk above

-   `activate()` of **sfdep** package is used to activate the geometry context.
-   `mutate()` of **dplyr** package is used to create two new columns *nb* and *wt*.
-   Then we activate the data context again and copy over the *nb* and *wt* columns to each time-slice using `set_nbs()` and `set_wts()`.
    -   Row order is very important, so do not rearrange the observations after using `set_nbs()` or `set_wts()`.
:::

Note that this dataset now has neighbours and weights for each time-slice.

### 5.2 Computing Gi\*

We can use these new columns to manually calculate the local Gi\* for each location. We do this by grouping by *Year* and using `local_gstar_perm()` of **sfdep** package. After which, we use `unnest()` to unnest the *gi_star* column of the newly created *gi_stars* data.frame.

```{r}
set.seed(1234)
gi_stars <- GDPPC_nb %>%
  group_by(Year) %>%
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>%
  tidyr::unnest(gi_star)
```

## 6 Mann-Kendall Test

A **monotonic series** or function is one that only increases (or decreases) and never changes direction. So long as the function either stays flat or continues to increase, it is monotonic.

-   H0: No monotonic trend
-   H1: Monotonic trend is present

**Interpretation**

-   Reject the null hypothesis if the p-value is smaller than the alpha value (i.e. 1 - confidence level).
-   Tau ranges between -1 and 1 where:
    -   -1 is a perfectly decreasing series, and
    -   1 is a perfectly increasing series.

### 6.1 Mann-Kendall test on Gi\*

With these Gi\* measures we can then evaluate each location for a trend using the Mann-Kendall test. The code chunk below uses Changsha county.

```{r}
cbg <- gi_stars %>%
  ungroup() %>%
  filter(County == "Changsha") %>%
  select(County, Year, gi_star)
```

Next, we plot the result by using **ggplot2** functions.

```{r}
ggplot(data = cbg,
       aes(x = Year,
           y = gi_star)) +
  geom_line() +
  theme_light()
```

### 6.2 Interactive Mann-Kendall plot

We can also create an interactive plot by using `ggplotly()` of **plotly** package.

```{r}
p <- ggplot(data = cbg,
            aes(x = Year,
                y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

### 6.3 Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
```

In the above result, **sl** is the p-value. If the p-value is smaller than 0.05, we reject the null hypothesis and infer that a monotonic trend is present; the sign of **tau** tells whether the trend is upward (positive) or downward (negative).

### 6.4 Mann-Kendall test data.frame

We can replicate this for each location by using `group_by()` of **dplyr** package.

```{r}
ehsa_mk <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa_mk)
```

### 6.5 Sorting to show significant emerging hot/cold spots

```{r}
emerging <- ehsa_mk %>%
  arrange(sl, abs(tau)) %>%
  slice(1:10)
head(emerging)
```

## 7 Performing Emerging Hotspot Analysis

Lastly, we perform EHSA by using `emerging_hotspot_analysis()` of **sfdep** package. It takes a spacetime object x (i.e. *GDPPC_st*), and the quoted name of the variable of interest (i.e. *GDPPC*) for the `.var` argument. The `k` argument is used to specify the number of time lags, which is set to 1 by default. Lastly, `nsim` defines the number of simulations to be performed.

```{r}
set.seed(1234)
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st,
  .var = "GDPPC",
  k = 1,
  nsim = 99
)
```

### 7.1 Visualising the distribution of EHSA classes

In the code chunk below, **ggplot2** functions are used to reveal the distribution of EHSA classes as a bar chart.

```{r}
#| fig-width: 10
#| fig-height: 6
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

### 7.2 Visualising EHSA

Before we can visualise the geographic distribution of EHSA classes, we need to join both *hunan* and *ehsa* together by using the code chunk below.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

Next, **tmap** functions are used to plot a categorical choropleth map. Only counties with statistically significant results (p-value \< 0.05) are coloured by their EHSA class.

```{r}
#| fig-width: 9
#| fig-height: 9
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```