---
title: "Hands-on Exercise 05: Spatio-Temporal Point Patterns Analysis"
author: "Xiaohan"
date: "today"
execute:
  eval: true
  echo: true
  freeze: true
---
#1 Overview
A spatio-temporal point process (also called space-time or spatial-temporal point process) is a random collection of points, where each point represents the time and location of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years. This chapter shows how various R packages can be combined to run a set of spatio-temporal point pattern analyses in a guided and intuitive way. A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

#2 Learning Outcome
The research questions
The specific questions we would like to answer are:

are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?
if the answer is NO, where and when the observed forest fire locations tend to cluster?

#3 The data
For the purpose of this exercise, two data sets will be used, they are:

forestfires, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from Fire Information for Resource Management System. For the purpose of this exercise, only forest fires within Kepulauan Bangka Belitung will be used.
Kepulauan_Bangka_Belitung, an ESRI shapefile showing the sub-district (i.e. kelurahan) boundary of Kepulauan Bangka Belitung. The data set was downloaded from Indonesia Geospatial portal. The original data covers the whole Indonesia. For the purpose of this exercise, only sub-districts within Kepulauan Bangka Belitung are extracted.

#4 Installing and Loading the R packages
```{r}
pacman::p_load(sf, spatstat, sparr, 
               tmap, stpp, tidyverse)
```

#5 Importing and Preparing Study Area
##5.1 Importing study area
Code chunk below is used import study area (i.e. Kepulauan Bangka Belitung) into R environment.

```{r}
kbb <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

The revised code chunk.

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

##5.2 Converting OWIN
Next, as.owin() is used to convert kbb into an owin object.

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

Next, class() is used to confirm if the output is indeed an owin object

```{r}
class(kbb_owin)
```

#6 Importing and Preparing Forest Fire data
Next, we will import the forest fire data set (i.e. forestfires.csv) into R environment.

```{r}
fire_sf <- read_csv("data/fire_archive_M-C61_802038.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326) %>%
  st_transform(crs = 32748)
```

Because ppp object only accept numerical or character as mark. The code chunk below is used to convert data type of acq_date to numeric.

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

#7 Visualising the Fire Points
##7.1 Overall plot

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

##7.2 Visuaising geographic distribution of forest fires by month

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

#8 Computing STKDE by Month
In this section, you will learn how to compute STKDE by using spattemp.density() of sparr package. Before using the function, it is highly recommended you read the function’s reference guide in detail in order to understand the input data requirements and the output object generated.

##8.1 Extracting forest fires by month
The code chunk below is used to remove the unwanted fields from fire_sf sf data.frame. This is because as.ppp() only need the mark field and geometry field from the input sf data.frame.

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

##8.2 Creating ppp
The code chunk below is used to derive a ppp object called fire_month from fire_month sf data.frame.

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

The code chunk below is used to check the output is in the correct object class.

```{r}
summary(fire_month_ppp)
```

Next, we will check if there are duplicated point events by using the code chunk below.

```{r}
any(duplicated(fire_month_ppp))
```

##8.3 Including Owin object
The code chunk below is used to combine origin_am_ppp and am_owin objects into one.

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

As a good practice, plot() is used to plot ff_owin so that we can examine the correctness of the output object.

```{r}
#| fig-width: 10
#| fig-height: 10
plot(fire_month_owin, xlim = c(512000, 750000), ylim = c(9700000, 9734000))
```

##8.4 Computing Spatio-temporal KDE
Next, spattemp.density() of sparr package is used to compute the STKDE.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

##8.5 Plotting the spatio-temporal KDE object
In the code chunk below, plot() of R base is used to the KDE for between July 2023 - December 2023.

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

#9 Computing STKDE by Day of Year
In this section, you will learn how to computer the STKDE of forest fires by day of year.

##9.1 Creating ppp object
In the code chunk below, DayofYear field is included in the output ppp object.

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

##9.2 Including Owin object
Next, code chunk below is used to combine the ppp object and the owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

##9.3

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```
```{r}
plot(kde_yday)
```

#10 Computing STKDE by Day of Year: Improved method
One of the nice function provides in sparr package is BOOT.spattemp(). It support bandwidth selection for standalone spatiotemporal density/intensity based on bootstrap estimation of the MISE, providing an isotropic scalar spatial bandwidth and a scalar temporal bandwidth.

Code chunk below uses BOOT.spattemp() to determine both the spatial bandwidth and the scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

##10.1 Computing spatio-temporal KDE
Now, the STKDE will be derived by using h and lambda values derive in previous step.

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

##10.2 Plotting the output spatio-temporal KDE
Last, plot() of sparr package is used to plot the output as shown below.

```{r}
plot(kde_yday)
```

#11 Spatio-temporal Point Patterns Analysis: stpp methods
In this section, you will gain hands-on experience on using functions of stpp package to perform spatio-temporal point patterns analysis.

Students are encouraged to read stpp: An R Package for Plotting, Simulating and Analyzing Spatio-Temporal Point Patterns to learn more about the package.

##11.1 Preparing spatio-temporal point process object of stpp
Step 1: Extracting forest fire coordinates from the fire point events

```{r}
coords <- st_coordinates(fire_sf)
```

Step 2: Creating a data frame by combining the x- and y-coordinates and temporal event. Note that the temporal event must be in integer.

```{r}
fire_df <- data.frame(
  x = coords[, 1],  
  y = coords[, 2],
  t = fire_sf$`DayofYear`)
```

Step 3: Creating stpp spatio-temporal object

In the code chunk below, as.3dpoint() of stpp package is used to create stpp spatio-temporal object class.

```{r}
fire_stpp <- as.3dpoints(fire_df)
```

Use the code chunk below to confirm that the output is in stpp spatio-temporal object class.

```{r}
class(fire_stpp)
```

Next we can visual fire_stpp by using the code chunk below.

```{r}
plot(fire_stpp)
```

##11.2 Computing spatio-temporal k-function
In the code chunk below, STIKhat() of stpp package is used to compute space-time inhomogeneous K-function.

```{r}
set.seed(123)
n_sample <- 10000
idx <- sample(nrow(fire_stpp), n_sample)
fire_stpp_small <- fire_stpp[idx, ]
class(fire_stpp_small) <- "stpp"

kbb_stik <- STIKhat(fire_stpp_small)
```

Next, plotK() is used to visualise the output space-time inhomogeneous K-function.

```{r}
plotK(kbb_stik)
```

##11.3 Guide to interpret the plot
If the contours show high values for small values of uu and vv, it suggests clustering at short spatial and temporal distances, meaning events occur close to each other in both space and time.
If the contours are flat or show low values, it indicates a more random distribution or a lack of significant clustering at those distances.
The spatial and temporal extent of clustering can be understood by observing how rapidly the contours increase or decrease as we move along the axes. In the plot above, we can see clustering at specific distances by observing the spacing and values of the contour lines.