---
title: "In-class_Ex04"
author: "Xiaohan Wang"
date: "september 18, 2026"
date-modified: today
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
editor: visual
---
```{r}
pacman::p_load(sf, ggstatsplot, tmap, spatstat, sparr, knitr, 
               tmap, stpp, tidyverse, GWmodel)
```

# preparing the data

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```
```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
hunan_sf <- left_join(hunan_sf, hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```
# Mapping GDPPC
```{r}
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```


```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
```
```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                 data = hunan_sp,
                 approach = "AICc",
                 adaptive = TRUE,
                 kernel = "bisquare",
                 longlat = T)
```


```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)
```

# prepaering the output data
Code chunk below is used to extract SDF data table from gwss object output from gwss() fro subsequent geovilisation. It will be converted into data.frame by using as.data.frame().
```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

Next,cbind() is used to append the newly derived data.frame onto hunan_sf sf data.frame.
```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

# visual geographically weighted summary statistic
```{r}
#| fig-width: 18
#| fig-height: 8
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of geographically wieghted mean") +
  tm_layout(legend.position = c("left", "bottom"))
```

