---
title: "Where Do Road Accidents Concentrate in the Bangkok Metropolitan Region?"
subtitle: "ISSS626 · Take-home Exercise 1 · Executive Summary"
author: "Wang Xiaohan"
date: "2026-09-27"
format:
  revealjs:
    theme: [default, summary.scss]
    width: 1280
    height: 720
    slide-number: true
    transition: fade
    scrollable: true
    footer: "Road traffic accidents in BMR, 2022"
execute:
  echo: false
  eval: true
  message: false
  warning: false
  freeze: true
---

```{r}
#| include: false
# Objects saved by the Technical Report (Take-home_Ex01.qmd) are reused here,
# so that the slides render quickly. Render the Technical Report first.
pacman::p_load(sf, spNetwork, tmap, tidyverse, spatstat, sparr)

bmr_names <- c("Bangkok", "Nonthaburi Province", "Nakhon Pathom Province",
               "Pathum Thani Province", "Samut Prakan Province",
               "Samut Sakhon Province")

bmr_sf <- st_read(dsn = "data/geospatial",
                  layer = "geoBoundaries-THA-ADM1",
                  quiet = TRUE) %>%
  filter(shapeName %in% bmr_names) %>%
  st_transform(crs = 32647)

bmr_union <- st_union(bmr_sf)
bmr_owin <- as.owin(bmr_union)

network <- read_rds("data/rds/network.rds")
acc_bmr_sf <- read_rds("data/rds/acc_bmr_sf.rds")
lixels <- read_rds("data/rds/lixels.rds")
kfun_acc <- read_rds("data/rds/kfun_acc.rds")
L_acc.csr <- read_rds("data/rds/L_acc_csr.rds")

# website colours
site_blue <- "#5eb3dd"
site_purple <- "#5e65ad"
site_accent <- "#d64482"
site_ramp <- c("#9fd3ee", "#5eb3dd", "#5e65ad", "#a44a93", "#d64482")
```

## Contents

::: {.toc-grid}
1.  Why this matters
2.  Study area and data
3.  Data quality and the observation process
4.  Analytical approach
5.  When do accidents happen?
6.  Where do accidents concentrate on the network?
7.  Are hotspots stable over the year?
8.  Are accidents clustered? It depends on the null model
9.  Limitations
10. Public-safety implications
:::

## 1. Why this matters

::: {.stats}
::: {.stat}
[18,218]{.num} [road deaths in Thailand, 2021 (WHO)]{.label}
:::

::: {.stat}
[25.4]{.num} [deaths per 100,000, among the highest in Asia]{.label}
:::

::: {.stat}
[3,599]{.num} [recorded accidents in BMR, 2022]{.label}
:::
:::

:::: {.columns}
::: {.column width="48%"}
::: {.card}
[The problem]{.eyebrow}

The Bangkok Metropolitan Region (BMR) is Thailand's largest urban area. But a concentration of accidents **is not the same as higher risk**: it may reflect traffic volume, road density or how accidents are recorded.
:::
:::

::: {.column width="52%"}
::: {.card}
[Research questions]{.eyebrow}

1.  **Where and when** are recorded accidents concentrated?
2.  Along the road network, are accidents **clustered, regular or random**?
3.  Do conclusions change between **planar and network** views?
:::
:::
::::

## 2. Study area and data

:::: {.columns}
::: {.column width="52%"}
```{r}
#| fig-width: 6
#| fig-height: 5
tm_shape(bmr_sf) +
  tm_polygons(fill = "grey95", col = "grey60") +
  tm_text("shapeName", size = 0.6) +
tm_shape(acc_bmr_sf) +
  tm_dots(size = 0.05, fill = site_accent) +
tm_title("3,599 recorded accidents, BMR 2022")
```
:::

::: {.column width="48%"}
::: {.card}
[Study area]{.eyebrow}

Bangkok + 5 vicinity provinces (Nonthaburi, Pathum Thani, Samut Prakan, Samut Sakhon, Nakhon Pathom), from geoBoundaries ADM1
:::

::: {.card}
[Data]{.eyebrow}

| Dataset | Source |
|----|----|
| Road accidents 2019-2022 (2022 used) | Kaggle |
| Province boundaries | geoBoundaries |
| Major roads (motorway to tertiary) | OpenStreetMap |
:::

::: {.takeaway}
**Point event:** one recorded accident in 2022, inside BMR, within 100 m of a major road
:::
:::
::::

## 3. Data quality and the observation process

:::: {.columns}
::: {.column width="50%"}
::: {.card}
[From raw records to events]{.eyebrow}

| Step | Records |
|----|---:|
| Raw records, 2019-2022 | 81,735 |
| Occurred in 2022 | 21,032 |
| Valid coordinates (189 missing removed) | 20,843 |
| Inside BMR | 3,599 |
| Within 100 m of a major road | 3,594 |

No swapped coordinates or duplicate records. Accidents sharing exact locations were kept.
:::
:::

::: {.column width="50%"}
::: {.card}
[Who recorded the accidents?]{.eyebrow}

| Agency | Share |
|----|---:|
| Department of Highways | 80.1% |
| Expressway Authority of Thailand | 13.1% |
| Department of Rural Roads | 6.8% |
| Bangkok Metropolitan Administration | 0% |
:::

::: {.takeaway}
All records come from **national road agencies**. Accidents on city-managed streets are likely missing, so results describe **recorded** accidents on highways, expressways and rural roads.
:::
:::
::::

## 4. Analytical approach

::: {.takeaway}
Accidents can only occur on roads, so the **road network is the study space**.
:::

::: {.card}
| Question | Main method | Reference / null model | Comparison |
|----|----|----|----|
| **When?** | Monthly and daily counts | – | Holiday campaign periods |
| **Where?** (first-order) | **Network KDE**, quartic kernel, 1 km along roads | – | Planar KDE; NKDE at 0.5 and 2 km |
| **Stable over time?** | Spatio-temporal KDE by day of year | – | STKDE by month |
| **Clustered?** (second-order) | **Network K- and G-functions**, 50 simulations, Bangkok | **CSR on the road network** | Clark-Evans, G and L vs planar CSR |
:::

::: {.card}
[Key assumptions]{.eyebrow}

Events lie on the major road network · network CSR treats every road as equally likely, ignoring traffic volume · recorded accidents stand in for all accidents (they do not, see slide 3)
:::

## 5. When do accidents happen?

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 5.5
#| fig-height: 3.4
ggplot(acc_bmr_sf, aes(x = Month_num)) +
  geom_histogram(binwidth = 1, fill = site_purple, colour = "white") +
  labs(x = "Month (1 = January)", y = "No. of accidents",
       title = "Monthly accidents, 2022") +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r}
#| fig-width: 5.5
#| fig-height: 3.4
ggplot(acc_bmr_sf, aes(x = DayofYear)) +
  geom_histogram(binwidth = 1, fill = site_blue) +
  geom_vline(xintercept = c(4, 101, 107, 363),
             linetype = "dashed", colour = site_accent, linewidth = 0.8) +
  labs(x = "Day of year", y = "No. of accidents per day",
       title = "Daily accidents (pink: holiday campaigns)") +
  theme_minimal()
```
:::
::::

:::: {.columns}
::: {.column width="50%"}
::: {.card}
Monthly counts range from about **210 (February)** to **395 (October)**, higher in the second half of 2022.
:::
:::

::: {.column width="50%"}
::: {.card}
The largest daily peaks coincide with the **New Year and Songkran** campaigns, up to about 36 a day.
:::
:::
::::

::: {.takeaway}
Counts are not adjusted for traffic volume: peaks mean **more recorded accidents**, not necessarily higher risk.
:::

## 6. Where do accidents concentrate on the network?

:::: {.columns}
::: {.column width="60%"}
```{r}
#| fig-width: 6.5
#| fig-height: 5.4
lixels_pos <- lixels %>%
  filter(density > 0)

tm_shape(bmr_sf) +
  tm_polygons(fill = "grey97", col = "grey70") +
tm_shape(network) +
  tm_lines(col = "grey80") +
tm_shape(lixels_pos) +
  tm_lines(col = "density",
           col.scale = tm_scale_intervals(n = 5,
                                          style = "quantile",
                                          values = site_ramp),
           col.legend = tm_legend(title = "NKDE (events per km)"),
           lwd = 2) +
tm_title("Network KDE, bandwidth 1 km")
```
:::

::: {.column width="40%"}
::: {.card}
[Finding]{.eyebrow}

Accidents concentrate on **a few specific roads**:

-   across **southern Bangkok into Samut Prakan** (highest density)
-   a **north–south corridor** into Pathum Thani
-   a **south-western corridor** towards Samut Sakhon
:::

::: {.takeaway}
Most of the major road network, including much of inner Bangkok, has **zero or near-zero** density. NKDE pinpoints road segments; planar KDE shows the same corridors but spreads them into areas without roads.
:::
:::
::::

## 7. Are hotspots stable over the year?

```{r}
#| include: false
acc_yday_ppp <- acc_bmr_sf %>%
  select(DayofYear) %>%
  as.ppp()
acc_yday_owin <- acc_yday_ppp[bmr_owin]

# same h and lambda as Section 6.6 of the Technical Report
kde_yday <- spattemp.density(acc_yday_owin,
                             h = 3062,
                             lambda = 76)
```

:::: {.columns}
::: {.column width="60%"}
```{r}
#| fig-width: 7
#| fig-height: 5.4
tims <- c(46, 105, 196, 288)
par(mfcol = c(2, 2))
for (i in tims) {
  plot(kde_yday, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("KDE at day", i))
}
```
:::

::: {.column width="40%"}
::: {.card}
[Stable core]{.eyebrow}

The hotspot in south-eastern Bangkok near Samut Prakan, and the corridor into Pathum Thani, **persist all year**.
:::

::: {.card}
[Changing periphery]{.eyebrow}

The corridor towards **Samut Sakhon** is faint in February and April, but prominent by **July and October**.
:::

::: {.takeaway}
Bootstrap bandwidths: 3.1 km, 76 days. This shows **seasonal** change; short holiday peaks are smoothed out (see slide 5).
:::
:::
::::

## 8. Are accidents clustered? It depends on the null model

:::: {.columns}
::: {.column width="50%"}
[Planar L-function vs planar CSR · BMR]{.eyebrow}

```{r}
#| fig-width: 5.5
#| fig-height: 3.6
plot(L_acc.csr, . - r ~ r,
     xlab = "d (km)", ylab = "L(d)-r",
     main = "")
```

::: {.card}
Strong clustering at all distances, but planar CSR assumes accidents can occur **anywhere**, including fields and rivers.
:::
:::

::: {.column width="50%"}
[Network K-function vs CSR on the network · Bangkok]{.eyebrow}

```{r}
#| fig-width: 5.5
#| fig-height: 3.6
kfun_acc$plotk +
  labs(x = "Network distance (m)") +
  theme_minimal()
```

::: {.card}
<!-- TODO: replace with the result from Section 7.1 of the Technical Report, e.g. "The observed curve lies above the envelope from X to Y m, indicating clustering along the network at these distances." -->
[Network K-function result]
:::
:::
::::

::: {.takeaway}
Much of the planar "clustering" reflects the fact that accidents are confined to a few roads. **The network test is the fairer comparison.**
:::

## 9. Limitations

:::: {.columns}
::: {.column width="50%"}
::: {.card}
[Data]{.eyebrow}

-   **Observation process:** only national road agencies' records; city streets are likely missing
-   **No exposure data:** without traffic volumes, concentration is not risk
-   **Location quality:** some accidents share exact locations, possibly fixed reference points
-   **Single year** (2022), and a **2026 road network** for 2022 accidents
:::
:::

::: {.column width="50%"}
::: {.card}
[Methods]{.eyebrow}

-   Results depend on **bandwidth**, tested at several values for planar KDE and NKDE
-   **Network CSR** assumes equal exposure on every road
-   Network K-function covers **Bangkok only**, with 50 simulations, for computational reasons
-   The day-of-year STKDE smooths out **short holiday peaks**
:::
:::
::::

## 10. Public-safety implications

:::: {.columns}
::: {.column width="50%"}
::: {.card}
[Supported directly by the analysis]{.eyebrow}

-   **Target road segments, not whole provinces** for audits, enforcement and engineering reviews
-   **Investigate repeated locations** at site level: junction design, signage, lighting
-   **Time campaigns** for New Year, Songkran and the second half of the year
-   **Monitor** the emerging Samut Sakhon corridor
-   **Check rescue-unit coverage** along the high-density corridors
:::
:::

::: {.column width="50%"}
::: {.card}
[Requires additional information]{.eyebrow}

-   **Traffic volumes**, to turn counts into risk per vehicle-kilometre
-   **Cause, vehicle and road-design data**, before designing interventions
-   **City-street records** from the Bangkok Metropolitan Administration
-   **Multi-year data**, to confirm patterns and evaluate campaigns
:::
:::
::::

::: {.takeaway}
These findings show where **recorded** accidents concentrate; they do not show why they occur.
:::
