---
title: "Where Do Road Accidents Concentrate in the Bangkok Metropolitan Region?"
subtitle: "Executive Summary · Take-home Exercise 1 · ISSS626 Geospatial Analytics"
author: "Wang Xiaohan"
date: "2026-09-27"
format:
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
    transition: fade
    scrollable: true
    footer: "Road traffic accidents in BMR, 2022 · Executive Summary"
execute:
  echo: false
  eval: true
  message: false
  warning: false
  freeze: true
---

```{r}
#| include: false
# Objects saved by the Technical Report (Take-home_Ex01.qmd) are reused here,
# so that the slides render quickly. Render the Technical Report first.
pacman::p_load(sf, spNetwork, tmap, tidyverse, spatstat, sparr)

bmr_names <- c("Bangkok", "Nonthaburi Province", "Nakhon Pathom Province",
               "Pathum Thani Province", "Samut Prakan Province",
               "Samut Sakhon Province")

bmr_sf <- st_read(dsn = "data/geospatial",
                  layer = "geoBoundaries-THA-ADM1",
                  quiet = TRUE) %>%
  filter(shapeName %in% bmr_names) %>%
  st_transform(crs = 32647)

bmr_union <- st_union(bmr_sf)
bmr_owin <- as.owin(bmr_union)

network <- read_rds("data/rds/network.rds")
acc_bmr_sf <- read_rds("data/rds/acc_bmr_sf.rds")
lixels <- read_rds("data/rds/lixels.rds")
kfun_acc <- read_rds("data/rds/kfun_acc.rds")
L_acc.csr <- read_rds("data/rds/L_acc_csr.rds")
```

## Contents

1.  Why this matters and research questions
2.  Study area and data
3.  Data quality and the observation process
4.  Analytical approach
5.  When do accidents happen?
6.  Where do accidents concentrate on the network?
7.  Are hotspots stable over the year?
8.  Are accidents clustered? It depends on the null model
9.  Limitations
10. Public-safety implications

## 1. Why this matters

:::: {.columns}
::: {.column width="50%"}
**The problem**

-   Thailand had about **18,218 road deaths in 2021**, a rate of **25.4 per 100,000**, among the highest in Asia (WHO, 2023)
-   The Bangkok Metropolitan Region (BMR) is Thailand's largest urban area
-   A concentration of accidents **is not the same as higher risk**: it may reflect traffic volume, road density or how accidents are recorded
:::

::: {.column width="50%"}
**Research questions**

1.  **Where and when** are recorded road accidents in BMR concentrated in 2022?
2.  Along the road network, are accidents **clustered, regular or random**, compared with random placement on the network?
3.  Do conclusions change between **planar and network** views?
:::
::::

## 2. Study area and data

:::: {.columns}
::: {.column width="55%"}
```{r}
#| fig-width: 6
#| fig-height: 5
tm_shape(bmr_sf) +
  tm_polygons(fill = "grey95") +
  tm_text("shapeName", size = 0.6) +
tm_shape(acc_bmr_sf) +
  tm_dots(size = 0.05, fill = "red") +
tm_title("3,599 recorded accidents, BMR 2022")
```
:::

::: {.column width="45%"}
**Study area:** Bangkok + 5 vicinity provinces (Nonthaburi, Pathum Thani, Samut Prakan, Samut Sakhon, Nakhon Pathom), from geoBoundaries ADM1

**Data**

| Dataset | Source |
|----|----|
| Road accidents 2019-2022 (2022 used) | Kaggle |
| Province boundaries | geoBoundaries |
| Major roads (motorway to tertiary) | OpenStreetMap via Geofabrik |

**Point event:** one recorded accident in 2022, inside BMR, within 100 m of a major road
:::
::::

## 3. Data quality and the observation process

:::: {.columns}
::: {.column width="50%"}
**From raw records to analysis-ready events**

| Step | Records |
|----|---:|
| Raw records, 2019-2022 | 81,735 |
| Occurred in 2022 | 21,032 |
| Valid coordinates (189 missing removed) | 20,843 |
| Inside BMR | 3,599 |
| Within 100 m of a major road | 3,594 |

No swapped coordinates or duplicate records were found. Some accidents share exact locations; they were kept.
:::

::: {.column width="50%"}
**Who recorded the accidents?**

| Agency | Share |
|----|---:|
| Department of Highways | 80.1% |
| Expressway Authority of Thailand | 13.1% |
| Department of Rural Roads | 6.8% |
| Bangkok Metropolitan Administration | 0% |

::: {.callout-important}
All records come from national road agencies. Accidents on city-managed streets are **likely missing**, so results describe recorded accidents on highways, expressways and rural roads.
:::
:::
::::

## 4. Analytical approach

Accidents can only occur on roads, so the **road network is the study space**.

| Question | Main method | Reference / null model | Comparison |
|----|----|----|----|
| **When?** | Monthly and daily counts | – | Holiday campaign periods |
| **Where?** (first-order) | **Network KDE** (quartic kernel, 1 km along roads) | – | Planar KDE; NKDE at 0.5 and 2 km |
| **Stable over time?** | Spatio-temporal KDE by day of year | – | STKDE by month |
| **Clustered?** (second-order) | **Network K- and G-functions**, 50 simulations, Bangkok | **CSR on the road network** | Clark-Evans, G and L vs planar CSR |

<br>

**Key assumptions:** events are located on the major road network; network CSR treats every road as equally likely, ignoring traffic volume; recorded accidents represent all accidents (they do not, see slide 3).

## 5. When do accidents happen?

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 5
#| fig-height: 3.5
ggplot(acc_bmr_sf, aes(x = Month_num)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Month (1 = January)", y = "No. of accidents",
       title = "Monthly accidents, 2022") +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r}
#| fig-width: 5
#| fig-height: 3.5
ggplot(acc_bmr_sf, aes(x = DayofYear)) +
  geom_histogram(binwidth = 1) +
  geom_vline(xintercept = c(4, 101, 107, 363),
             linetype = "dashed", colour = "red") +
  labs(x = "Day of year", y = "No. of accidents per day",
       title = "Daily accidents (red: holiday campaigns)") +
  theme_minimal()
```
:::
::::

-   Monthly counts range from about **210 (February) to 395 (October)**, higher in the second half of 2022
-   The largest daily peaks coincide with the **New Year and Songkran** campaign periods, up to about 36 accidents a day
-   Counts are not adjusted for traffic volume, so peaks mean **more recorded accidents**, not necessarily higher risk

## 6. Where do accidents concentrate on the network?

:::: {.columns}
::: {.column width="62%"}
```{r}
#| fig-width: 6.5
#| fig-height: 5.5
lixels_pos <- lixels %>%
  filter(density > 0)

tm_shape(bmr_sf) +
  tm_polygons(fill = "grey95") +
tm_shape(network) +
  tm_lines(col = "grey80") +
tm_shape(lixels_pos) +
  tm_lines(col = "density",
           col.scale = tm_scale_intervals(style = "quantile",
                                          values = "viridis"),
           col.legend = tm_legend(title = "NKDE (events per km)"),
           lwd = 2) +
tm_title("Network KDE, bandwidth 1 km")
```
:::

::: {.column width="38%"}
**Accidents concentrate on a few specific roads**

-   Corridors across **southern Bangkok into Samut Prakan** (highest density)
-   A **north–south corridor** into Pathum Thani
-   A **south-western corridor** towards Samut Sakhon

Most of the major road network, including much of inner Bangkok, has **zero or near-zero** density.

Planar KDE shows the same corridors but spreads them into areas without roads; NKDE pinpoints the **road segments**.
:::
::::

## 7. Are hotspots stable over the year?

```{r}
#| include: false
acc_yday_ppp <- acc_bmr_sf %>%
  select(DayofYear) %>%
  as.ppp()
acc_yday_owin <- acc_yday_ppp[bmr_owin]

# same h and lambda as Section 6.6 of the Technical Report
kde_yday <- spattemp.density(acc_yday_owin,
                             h = 3062,
                             lambda = 76)
```

:::: {.columns}
::: {.column width="62%"}
```{r}
#| fig-width: 7
#| fig-height: 5.5
tims <- c(46, 105, 196, 288)
par(mfcol = c(2, 2))
for (i in tims) {
  plot(kde_yday, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("KDE at day", i))
}
```
:::

::: {.column width="38%"}
**Spatio-temporal KDE** (bootstrap bandwidths: 3.1 km, 76 days)

-   **Stable core:** the hotspot in south-eastern Bangkok near Samut Prakan, and the corridor into Pathum Thani, persist all year
-   **Changing periphery:** the corridor towards **Samut Sakhon** is faint in February and April, but prominent by July and October

The 76-day bandwidth shows **seasonal** change; short holiday peaks are smoothed out (see slide 5).
:::
::::

## 8. Are accidents clustered? It depends on the null model

:::: {.columns}
::: {.column width="50%"}
**Planar L-function vs planar CSR (BMR)**

```{r}
#| fig-width: 5
#| fig-height: 4
plot(L_acc.csr, . - r ~ r,
     xlab = "d (km)", ylab = "L(d)-r",
     main = "")
```

Strong clustering at all distances, but planar CSR assumes accidents can occur **anywhere**, including fields and rivers.
:::

::: {.column width="50%"}
**Network K-function vs CSR on the network (Bangkok)**

```{r}
#| fig-width: 5
#| fig-height: 4
kfun_acc$plotk +
  labs(x = "Network distance (m)")
```

<!-- TODO: replace with the result from Section 7.1 of the Technical Report, e.g. "The blue line lies above the envelope from X to Y m, indicating clustering along the network at these distances." -->
[Network K-function result]
:::
::::

**Takeaway:** much of the planar "clustering" reflects the fact that accidents are confined to a few roads. The network test is the fairer comparison.

## 9. Limitations

:::: {.columns}
::: {.column width="50%"}
**Data**

-   **Observation process:** only national road agencies' records; city streets are likely missing
-   **No exposure data:** without traffic volumes, concentration ≠ risk
-   **Location quality:** some accidents share exact locations, possibly fixed reference points
-   **Single year** (2022) and a **2026 road network** for 2022 accidents
:::

::: {.column width="50%"}
**Methods**

-   Results depend on **bandwidth** (tested at several values for planar KDE and NKDE)
-   **Network CSR** assumes equal exposure on every road
-   Network K-function covers **Bangkok only**, with 50 simulations, for computational reasons
-   The day-of-year STKDE smooths out **short holiday peaks**
:::
::::

## 10. Public-safety implications

:::: {.columns}
::: {.column width="50%"}
**Supported directly by the analysis**

-   **Target road segments, not whole provinces:** focus audits, enforcement and engineering reviews on the high-density corridors
-   **Investigate repeated locations** at site level (junction design, signage, lighting)
-   **Time campaigns** for the New Year and Songkran periods and the second half of the year
-   **Monitor** the emerging Samut Sakhon corridor
-   **Emergency response:** check rescue-unit coverage along the high-density corridors
:::

::: {.column width="50%"}
**Requires additional information**

-   **Traffic volumes**, to turn counts into risk per vehicle-kilometre
-   **Cause, vehicle and road-design data**, before designing specific interventions
-   **City-street records** from the Bangkok Metropolitan Administration
-   **Multi-year data**, to confirm that the patterns persist and to evaluate campaigns
:::
::::

::: {.callout-note}
These findings describe where **recorded** accidents concentrate; they do not show why they occur.
:::
