---
title: "Hands-on Exercise 01: Geospatial Data Wrangling"
author: "Wang Xiaohan"
date: "`r Sys.Date()`"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
editor: visual
---

## 1 Getting Started What we can learn from the code chunk above:

p_load function pf pacman package is used to install and load sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## 2 Importing Geospatial Data

In this section, you will learn how to import the following geospatial data into R by using `st_read()` of **sf** package:

- `MasterPlan2014SubzoneBoundaryNoSea`, a polygon feature layer in geojson format,
- `CyclingPathGazette`, a line feature layer in ESRI shapefile format, and
- `PreSchoolsLocation`, a point feature layer in geojson format.

### 2.1 Importing polygon feature data

The code chunk below uses `st_read()` function of **sf** package to import `MasterPlan2014SubzoneBoundaryNoSea` geojson file into R as a polygon feature data frame.

```{r}
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

The message above reveals that the geospatial objects are multipolygon features. `mpsz` is in svy21 projected coordinates system. The bounding box provides the x extend and y extend of the data.

### 2.2 Importing polyline feature data

The code chunk below uses `st_read()` function of **sf** package to import `CyclingPathGazette` shapefile into R as line feature data frame. Note that when the input geospatial data is in shapefile format, two arguments will be used, namely: `dsn` to define the data path and `layer` to provide the shapefile name.

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                       layer = "CyclingPathGazette")
```

The message above reveals that the geospatial entities are captured in multilinestring object. This data set is in svy21 projected coordinates system too.

### 2.3 Importing point feature data

The code chunk below will be used to import `PreSchoolsLocation` geojson file into R. Since the coordinate system of this data set is wgs84, `st_transform()` is used to project it into svy21 projected coordinates system.

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.geojson") %>%
  st_transform(crs = 3414)
```

## 3 Checking the Content of A Simple Feature Data Frame

Prior to wrangle the geospatial data, it is always a good practice for us to know more about the data sets first.

### 3.1 Working with st_geometry()

The column in the sf data.frame that contains the geometries is a list of class `sfc`. We can retrieve the geometry list-column by using `st_geometry()` as shown in the code chunk below.

```{r}
st_geometry(mpsz)
```

Notice that the print only displays basic information of the feature class such as type of geometry, the geographic extent of the features and the coordinate system of the data.

### 3.2 Working with glimpse()

Beside the basic geospatial feature information, we also would like to learn more about the associated attribute information in the data frame. This is the time you will find `glimpse()` of **dplyr** very handy as shown in the code chunk below.

```{r}
glimpse(mpsz)
```

`glimpse()` report reveals the data type of each field.

### 3.3 Working with head()

Sometimes we would like to reveal complete information of a feature object, this is the job of `head()` of Base R.

```{r}
head(mpsz, n = 5)
```

One of the useful arguments of `head()` is it allows user to select the numbers of record to display (i.e. the n argument).

## 4 Plotting the Geospatial Data

In geospatial data science, looking at the feature information is not enough. We are also interested to visualise the geospatial features. This is the time you will find `plot()` of R Graphic comes in very handy as shown in the code chunk below.

```{r}
plot(mpsz)
```

The default plot of an sf object is a multi-plot of all attributes, up to a reasonable maximum. We can, however, choose to plot only the geometry by using the code chunk below.

```{r}
plot(st_geometry(mpsz))
```

Alternatively, we can also choose to plot the sf object by using a specific attribute as shown in the code chunk below.

```{r}
plot(mpsz["PLN_AREA_N"])
```

`plot()` is appropriate for plotting the geospatial object for quick look. For high cartographic quality map, other R package such as **tmap** should be used.

Now, let us plot the preschool layer ontop of the mpsz layer by using the code chunk below.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

Note that the preschool point objects may fail to plot ontop of the mpsz layer if the coordinate systems do not match. Do you know why?

## 5 Working with Projection

Map projection is an important property of a geospatial data. In order to perform geoprocessing using two geospatial data, we need to ensure that both geospatial data are projected using similar coordinate system.

### 5.1 Assigning EPSG code to a simple feature data frame

One of the common issues that can happen during importing geospatial data into R is that the coordinate system of the source data was either missing or wrongly assigned during the importing process.

This is an example the coordinate system of `mpsz` simple feature data frame by using `st_crs()` of **sf** package as shown in the code chunk below.

```{r}
st_crs(mpsz)
```

If the print reveals a wrong EPSG code, we need to assign the correct EPSG code to `mpsz` data frame, `st_set_crs()` of **sf** package is used as shown in the code chunk below.


```{r}
mpsz <- st_set_crs(mpsz, NA)
mpsz <- st_set_crs(mpsz, 4326)
mpsz <- st_transform(mpsz, 3414)
```

```{r}
st_bbox(mpsz)
```

Now, let us check the CRS again by using the code chunk below.

```{r}
st_crs(mpsz)
```

Notice that the EPSG code is 3414 now.

### 5.2 Transforming the projection of preschool from wgs84 to svy21

In geospatial analytics, it is very common for us to transform the original data from geographic coordinate system to projected coordinate system. This is because geographic coordinate system is not appropriate if the analysis need to use distance or/and area measurements.

This is a scenario that `st_set_crs()` is not appropriate and `st_transform()` of sf package should be used. This is because we need to reproject preschool from one coordinate system to another coordinate system mathematically. This was already performed in Section 2.3 above.

Now, let us try to plot the preschool layer ontop of mpsz layer again by using the similar code chunk you used earlier.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

Notice that the composite map displays like what we want now.

## 6 Importing and Converting An Aspatial Data

In practice, it is not unusual that we will come across data such as `listing` of Inside Airbnb. We call this kind of data aspatial data. This is because it is not a geospatial data but among the data fields, there are two fields that capture the x- and y-coordinates of the data points.

In this section, you will learn how to import an aspatial data into R environment and save it as a tibble data frame. Next, you will convert it into a simple feature data frame.

### 6.1 Importing the aspatial data

Since `listings` data set is in csv file format, we will use `read_csv()` of **readr** package to import `listings.csv` as shown the code chunk below. The output R object is called `listings` and it is a tibble data frame.

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

After importing the data file into R, it is important for us to examine if the data file has been imported correctly.

The code chunk below shows `list()` of Base R instead of `glimpse()` is used to do the job.

```{r}
list(listings)
```

Two useful fields we are going to use in the next phase are `latitude` and `longitude`. Note that they are in decimal degree format. As a best guess, we will assume that the data is in wgs84 Geographic Coordinate System.

### 6.2 Creating a simple feature data frame from an aspatial data frame

The code chunk below converts `listing` data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages.

```{r}
listings_sf <- st_as_sf(listings,
                         coords = c("longitude", "latitude"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

Things to learn from the arguments above:

- `coords` argument requires you to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.
- `crs` argument requires you to provide the coordinates system in epsg format. EPSG: 4326 is wgs84 Geographic Coordinate System and EPSG: 3414 is Singapore SVY21 Projected Coordinate System.
- `%>%` is used to nest `st_transform()` to transform the newly created simple feature data frame into svy21 projected coordinates system.

Let us examine the content of this newly created simple feature data frame.

```{r}
glimpse(listings_sf)
```

Table above shows the content of `listings_sf`. Notice that a new column called `geometry` has been added into the data frame. On the other hand, the `longitude` and `latitude` columns have been dropped from the data frame.

**DIY**: Using the method you learned earlier, plot listings_sf layer on top of mpsz layer.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

## 7 Geoprocessing with sf package

Besides providing functions to handling (i.e. importing, exporting, assigning projection, transforming projection etc) geospatial data, **sf** package also offers a wide range of geoprocessing (also known as GIS analysis) functions.

In this section, you will learn how to answer GIS questions by using geoprocessing functions of sf package.

### 7.1 Use case 1: Land acquisition analysis

**The scenario**

The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 metres of reserved land on the both sides of the existing cycling path. You are tasked to determine the extend of the land need to be acquired and their total area.

**The solution**

Firstly, `st_buffer()` of **sf** package is used to compute the 5-meter buffers around cycling paths.

```{r}
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

This is followed by calculating the area of the buffers as shown in the code chunk below.

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

If you are tidyverse person like me, code chunk below should be used.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

Lastly, `sum()` of Base R will be used to derive the total land involved.

```{r}
sum(buffer_cycling$AREA)
```

Firstly, filter() of dplyr package will be used to extract polygon feature of Tampines West by using the code chunk below.
```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST") 
```

Next, st_intersection() of sf package will be used to clip cycling buffers within Tampines West planning subzone.
```{r}
buffer_cycling <- st_transform(buffer_cycling, 3414)
mpsz_selected <- st_transform(mpsz_selected, 3414)

buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```
```{r}
plot(st_geometry(buffer_cycling_selected))
```

Mission Accomplished!

### 7.2 Use case 2: To determine the number of pre-schools by planning subzone

**The scenario**

The authority requires a count of pre-schools for each planning subzone to support forward planning. Using R and the sf package, perform the necessary geoprocessing to compute these counts and present the results clearly.

**The solution**

The code chunk below performs two operations at one go. Firstly, identify pre-schools located inside each Planning Subzone by using `st_intersects()`. Next, `length()` of Base R is used to calculate numbers of pre-schools that fall inside each planning subzone.

```{r}
st_crs(mpsz)
st_crs(preschool)
```
```{r}
st_bbox(mpsz)
st_bbox(preschool)
```
```{r}
st_geometry(preschool)
```


```{r}
mpsz$`PreSch Count`<- lengths(st_intersects(mpsz, preschool))
```

You should not confuse with `st_intersection()`.

You can check the summary statistics of the newly derived `PreSch Count` field by using `summary()` as shown in the code chunk below.

```{r}
summary(mpsz$`PreSch Count`)
```

To list the planning subzone with the most number of pre-school, the `top_n()` of **dplyr** package is used as shown in the code chunk below.

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

In the code chunk below, another geoprocessing function of sf package called `st_area()` is used to derive the area of each planning subzone.

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

Next, `mutate()` of **dplyr** package is used to compute the density by using the code chunk below.

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## 8 Exploratory Data Analysis (EDA)

In this section, you will visualise the derive variables by using appropriate Exploratory data Analysis methods of **ggplot2**.

Firstly, we will plot a histogram to reveal the distribution of `PreSch Density`. Conventionally, `hist()` of R Graphics will be used as shown in the code chunk below.

```{r}
hist(mpsz$`PreSch Density`)
```

Although the syntax is very easy to use however the output is far from meeting publication quality. Furthermore, the function has limited room for further customisation.

In the code chunk below, appropriate **ggplot2** functions will be used.

```{r}
ggplot(data = mpsz,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                  color = "black",
                  fill = "light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle = "There are many planning subzones with a single pre-school, on the other hand, there are seven planning subzones with at least 30 or more pre-schools",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```

In the code chunk below, appropriate ggplot2 functions are used to plot a scatterplot showing the relationship between Pre-school Density and Pre-school Count.

```{r}
ggplot(data = mpsz,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black",
             fill = "light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
       x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```